---
title: "Finding the Story: Transportation Data"
author: "Cat Murphy"
output: html_notebook
---

# Finding the Story: Transportation Data

## Install & Load Packages

```{r install_packages}
# install.packages("tidyverse")
library(tidyverse)

# install.packages("dplyr")
library(dplyr)

# install.packages("janitor")
library(janitor)

# install.packages("lubridate")
library(lubridate)
```

## Load Rail Safety Recommendations (2000-present)

*Source: National Transportation Safety Board's CAROL Database*

*Link to Custom Search Builder: <https://data.ntsb.gov/carol-main-public/query-builder>*

The first step in our data exploration process is to load the raw .csv data from the NTSB's database to get a sense of what we're working with.

Note that this is *summary* data from NTSB's CAROL database, which means it's missing some of the contextual information that you can see online — like a recommendation's priority level, hazmat status, etc.

The reason? The full data is in a nested JSON file that by itself is pretty unmanageable. We'll come back to this, but for now, the summary data is what we'll focus on.

```{r load_data}
# load raw csv
raw_rail_recs <- read_csv("data/rail_safety_recs.csv") |>
  
  # clean names
  clean_names() |>
  
  # peek at data
  glimpse()
```

OK, so `glimpse()` shows us that we've got 21 columns. 

Most of the column names are pretty self-explanatory, but not all of them. Let's fix that.

## Initial Column Cleaning

```{r cleaning_and_renaming}
clean_rail_recs <- raw_rail_recs |>
  
  # rename columns
  rename(recommendation_id = srid,
         summary = abstract,
         recommendation_encodings = sr_coding_mode,
         recommendation_category = sr_coding_tier_1,
         recommendation_subcategory = sr_coding_tier2,
         recommendation_url = sr_url,
         recommendation_date_issued = date_issued,
         ) |>

  # standardize all text to uppercase
  mutate(across(everything(), ~str_to_upper(.)))
```

```{r separate_recommendations_by_addressee}
clean_rail_recs <- clean_rail_recs |>
  
  # remove trailing semicolons in recipient details
  mutate(addressee_details = str_remove_all(addressee_details, ";$")) |>

  # use remaining semicolons to put assign each recipient its own recommendation row when a single recommendation has multiple recipients
  separate_rows(addressee_details, sep = ";") |>
  
  # remove extra white space from addressee details
  mutate(addressee_details = str_squish(addressee_details))
```

```{r separate_addressee_details}
clean_rail_recs <- clean_rail_recs |>
  # separate recipient details into three distinct columns
  separate(addressee_details, 
           into = c("recipient", 
                    "recommendation_status", 
                    "recommendation_date_closed"), 
           sep = " \\| ")
```

```{r}
 # since some recommendations have multiple recipients with different statuses, we need to parse the recipient status column to get a more accurate picture of which recommendations are open vs. closed for each recipient

clean_rail_recs <- clean_rail_recs |>
  mutate(
    
    # step 1 — create open/closed column using case_when()
    recommendation_open_closed = case_when(
      str_detect(recommendation_status, "CLOSED") ~ "CLOSED",
      str_detect(recommendation_status, "OPEN") ~ "OPEN"),
    
    # step 2 — response status column using str_remove_all() & str_trim()
    recommendation_response_status = str_trim(str_remove_all(recommendation_status, "CLOSED|OPEN|-")))
```

```{r}
# drop original overall_status and recommendation_status columns since we've parsed out the relevant information into new columns!
clean_rail_recs <- clean_rail_recs |>
  select(-overall_status, -recommendation_status)
```

```{r date_handling}
# convert date columns to date format using as.Date() and mdy() from lubridate package (since some of the date formats are inconsistent across columns)

clean_rail_recs <- clean_rail_recs |>
  mutate(
    # convert overall date columns to date format using as.Date() since they are in a consistent format
    across(c(event_date, recommendation_date_issued), ~as.Date(.)),
    
    # convert recipient date column to date format using mdy() since it was extracted from the addressee_details column and is in a different format than the overall date columns
   recommendation_date_closed = mdy(recommendation_date_closed),
    
    # add year columns
    event_year = year(event_date),
    recommendation_year_issued = year(recommendation_date_issued),
    recommendation_year_closed = year(recommendation_date_closed)
  )

# *note: you will get a warning for the recipient_date_closed column. that's ok! the columns it's failing to convert are the ones that have NA values in them, which is expected since not all recommendations are closed
```

```{r}
# reorganize columns
clean_rail_recs <- clean_rail_recs |>
  select(recommendation_id, 
         ntsb_no,
         mkey,
         report_no,
         recommendation_mode,
         summary,
         event_date,
         event_year,
         city,
         state,
         country,
         recommendation,
         recommendation_encodings,
         recommendation_category,
         recommendation_subcategory,
         recommendation_date_issued,
         recommendation_year_issued,
         recipient,
         is_multiple_recipient,
         recommendation_response_status,
         recommendation_open_closed,
         recommendation_date_closed,
         recommendation_year_closed,
         recommendation_url,
         accident_report_url
         )
```

```{r}
json_data <- read_csv("data/rail_safety_json_data.csv")
```

```{r}
x <- clean_rail_recs |>
  left_join(json_data, join_by(recommendation_id, recommendation))
```
