---
title: "Finding the Story: Transportation Data"
author: "Code by Cat Murphy, Howard Center for Investigative Journalism at the University of Maryland"
output: html_notebook
---

## Step 1: Install & Load Packages

These are the packages we'll be using for this data exploration process. If you don't have them installed, you can run the install.packages() lines to install them. Otherwise, just load the libraries and we'll get going!

```{r install_packages}
# install.packages("tidyverse")
library(tidyverse)

# install.packages("dplyr")
library(dplyr)

# install.packages("janitor")
library(janitor)
```

## Step 2: Load Data & Clean Column Names

*Source: National Transportation Safety Board's CAROL Database*

*Link to Custom Search Builder: <https://data.ntsb.gov/carol-main-public/query-builder>*

### Use `load_csv()` to load the data

The first step in our data exploration process is to load that data we were working with.

We're going to use the `read_csv()` function to help us do that and then we'll clean the column names and peek at the data using `glimpse()` to get a sense of what we're working with.

```{r load_data}
# syntax:
# what_you_want_to_name_your_data <- read_csv("path/to/your/data.csv")

recs <- read_csv("data/rail_safety_recs.csv") 

# take a look to see what we have
glimpse(recs)
```

Great! But those column names are a little hard to work with.

### Use `clean_names()` to clean column names

So let's use the `clean_names()` function from the `janitor` package to make our column names easier to work with, and then we can use `glimpse()` to peek at the data again.

```{r clean_names_and_glimpse}
recs <- recs |>
  # clean names
  clean_names() |>
  
  # peek at data
  glimpse()
```

Alright, now that's much easier to work with!

## Step 3: Exploring & Cleaning the Data

### Explore recommendations with multiple recipients

OK, remember that example where one recommendation had multiple recipients with different statuses, and everything was jammed into one column? Let's check that out now that we have the data loaded and cleaned.

```{r}
recs |>
  # filter for recommendations with multiple recipients
  filter(is_multiple_recipient == TRUE) |>
  
  # select only the addressee details column to see what the recipient details look like for recommendations with multiple recipients
  select(addressee_details) |>
  
  # take a peek at the first five rows
  head(5)
```

#### **BIG YIKES.**

Like ... gross. How am I supposed to be able to look at who the NTSB is giving recommendations to and how — or if — they've responded when all of that information is in one messy column?

### What we know about the data

We can see that the `addressee_details` column contains:

-   The name of each organization or agency that was sent the recommendation

-   The statuses of the recommendation (open vs. closed)

-   How the NTSB has classified their response to the recommendation (acceptable vs. unacceptable, etc.)

-   The date the recommendation was closed (if applicable)

Of course, all of that info is super useful, but it's impossible to analyze it when it's all jammed together in one column.

What we we need to do is **clean up that column and separate it into distinct columns** for recipient name, recommendation status, and recommendation date closed.

If you look at the `addressee_details` column, you can see two things:

-   The recipients are separated by semicolons (;)

-   The different pieces of information are separated by a pipe character (\|)

We can use those characters to help us separate the information.

### Creating distinct rows for each recipient

Let's start with the semicolons. If you look close, in addition to the semicolons separating each recipient, there are actually trailing semicolons in the `addressee_details` column. So if we separate them as is, we'll end up with some rows without recipients because it will think that last semicolon means it should create a row. 

So let's clean those up first using a function called `str_remove_all` — literally, remove all instances of "the thing i want removed" from column_im_telling_you_to_remove_it_from. If you're newer to R, note that this is a function that can't stand on its own. You'll have to wrap it inside a `mutate` function so that R knows to, well, mutate a specific column. 

#### Removing trailing semicolons

```{r remove_trailing_semicolons}
recs <- recs |>
  mutate(addressee_details = str_remove_all(addressee_details, ";$"))

  # NOTE: the $ in the syntax is telling R that we only want to remove semicolons that are at the end of a string; not including it would mean removing all of the semicolons, which we don't want to do because we need them to separate the recipients into distinct rows in the next step

  # by saying we want addressee_details to equal the result of str_remove_all, we're telling R to replace the existing column with the new version. if we wanted a new column — let's call it flooberdoof — with the mutated version of addressee_details, we could say:
    # mutate(flooberdoof = str_remove_all(addressee_details, ";$"))

recs |>
  select(addressee_details) |>
  head(5)
```

No more trailing semicolons! But what are those weird newline elements? Those are just the way the data is formatted in the original source. Luckily, we can use `str_squish()` to fix that.

#### Removing extra white space with and newline characters with `str_squish()`

`str_squish()` — the most adorable-sounding function — just removes extra white space from the start and end of a string, and replaces all internal white spaces with a single space, so it's perfect for this job (not to mention awesome to say).

```{r}
recs <- recs |>
  # remove extra white space from addressee details
  mutate(addressee_details = str_squish(addressee_details))

# and let's take a look at the addressee_details column one more time to make sure the extra white space is gone!
recs |>
  filter(is_multiple_recipient == TRUE) |>
  select(addressee_details) |>
  head(5)
```

#### Using `separate_rows()` to create distinct rows for each recipient

Now that it looks more normal ... well, normal-ish ... we can use the remaining semicolons to separate the recipients into distinct rows. The creatively named `separate_rows()` function is perfect for the job. It does exactly what you think it does: it separates rows based on a separator that you specify. So in this case, we want to separate the rows in the `addressee_details` column based on the semicolons.

```{r separate_recommendations_by_addressee}
recs <- recs |>
  # use semicolons to put assign each recipient its own recommendation row when a single recommendation has multiple recipients
  separate_rows(addressee_details, sep = ";")

# let's take a peek at our data frame again
recs |>
  filter(is_multiple_recipient == TRUE) |>
  select(addressee_details) |>
  head(5)
```

Perfect! We have one row per recipient and no extra white space in the `addressee_details` column. 

### Creating separate columns for the information in the `addressee_details` column

Alright, but we still have all of the information for that recipient in the `addressee_details` column, which is pretty messy.

Let's use the pipe characters and the function `separate()` — which is very similar to `separate_rows()` — to separate that information into three columns: recipient name, recommendation status and the date the recommendation was closed (if applicable).

```{r separate_addressee_details}
recs <- recs |>
  # separate recipient details into three distinct columns
  separate(addressee_details, 
           into = c("recipient", 
                    "recipient_recommendation_status", 
                    "recipient_recommendation_date_closed"), 
           sep = " \\| ")


# all we're doing here is telling R to:
  # look at the addressee_details column
  # separate it into three new columns called recipient, recommendation_status and recommendation_date_closed
  # use the pipe character (|) as the separator

# we also have to add a space before and after the pipe character in the separator argument since there is a space before and after the pipe character in the addressee_details column.


# then we can check out the new columns we just created to make sure they look right!
recs |>
  filter(is_multiple_recipient == TRUE) |>
  select(recipient, recipient_recommendation_status, recipient_recommendation_date_closed) |>
  head(5)
```

Look at that! Now we have separate columns for recipient name, recommendation status and the date for when the recommendation was closed (if applicable).

#### Removing unnecessary columns

And because we have the recipient-specific columns now, we actually don't need the `overall_date_closed` or `overall_status` column anymore, so we can drop that using `select()` with a minus sign to essentially tell R to *deselect* those columns.

```{r}
recs <- recs |>
  # drop overall status and date closed columns since we now have recipient-specific status and date closed columns
  select(-overall_status, -overall_date_closed)
```

#### Transforming the newly created date column to date format

Because we extracted the new date column from the `addressee_details` column, it's in character format instead of date format, which means we won't be able to do any date calculations with it. The date is in a mm/dd/yyyy format, so we'll adjust it using the `mdy()` function. This is another function that needs to be put within a `mutate()` statement.

```{r}
recs <- recs |>
  # convert recipient date column to date format using mdy() since it was extracted from the addressee_details column and is in a different format than the overall date columns
  mutate(recipient_recommendation_date_closed = mdy(recipient_recommendation_date_closed))

recs |>
  select(recipient_recommendation_date_closed) |>
  tail(5)
```

Honestly, this is already a much more usable dataset than what we started with, but we can make it even better by parsing the recipient recommendation status column to get a more accurate picture of which whether each recipient's recommendation is open or closed and how the NTSB has classified their response to the recommendation.

Here, we're going to use `case_when()` to analyze the current column and create a new column that just tells us whether the recommendation is open or closed.

All a `case_when()` does is parse one column and create a new column based on whether certain conditions are met. In this case, the conditions are whether the `recipient_recommendation_status` column contains the word "CLOSED" or "OPEN", so we're going to use `str_detect()` — literally, detect "what i want to look for" in column_im_asking_you_to_look_for_it_in — to check for those words in the column and then create a new column called `recipient_recommendation_open_closed` that just tells us whether the recommendation is open or closed for each recipient.

```{r}
recs <- recs |>
  # step 1 — create open/closed column using case_when()
  mutate(
    recipient_recommendation_open_closed = case_when(
      
      str_detect(recipient_recommendation_status, "Closed") ~ "Closed",
      
      str_detect(recipient_recommendation_status, "Open") ~ "Open")
    )

recs |>
  filter(is_multiple_recipient == TRUE) |>
  select(recipient, recipient_recommendation_open_closed, recipient_recommendation_status, recipient_recommendation_date_closed) |>
  head(5)
```

Now we'll use `str_remove_all()`to get rid of those details, since we just put them in a new column called `recipient_recommendation_open_closed`. That way create another column that *just* tells us how the NTSB has classified the recipient's response to the recommendation (acceptable, unacceptable, etc.) without all of the extra information about whether it's open or closed. 

```{r}
recs <- recs |>  
  mutate(
    recipient_recommendation_response_status = 
      str_remove_all(recipient_recommendation_status, "Closed|Open|-")
    )


# here, the pipe (|) in the syntax means "and." it's just telling R to remove all instances of "Closed" AND "Open" AND "-" from the recipient_recommendation_status column and put the result in a new column called recipient_recommendation_response_status. it's a shorter way of writing this:
  # mutate(recipient_recommendation_response_status = str_remove_all(recipient_recommendation_status, "Closed"),
  #        recipient_recommendation_response_status = str_remove_all(recipient_recommendation_response_status, "Open"),
  #        recipient_recommendation_response_status = str_remove_all(recipient_recommendation_response_status, "-"))


recs |>
  filter(is_multiple_recipient == TRUE) |>
  select(recipient, recipient_recommendation_open_closed, recipient_recommendation_response_status, recipient_recommendation_date_closed) |>
  tail(5)
```

We can also use `str_squish()` just like we did earlier to clean up any extra white space that might be left over.

```{r remove_extra_white_space_from_response_status}
recs <- recs |>
  # remove extra white space from new recipient response status column
  mutate(recipient_recommendation_response_status = str_squish(recipient_recommendation_response_status))
```

Since we now have separate columns for whether the recommendation is open or closed and how the NTSB has classified the recipient's response to the recommendation, we can drop the original `recipient_recommendation_status` column since it contains all of that information in one messy string.

```{r drop_original_recommendation_status_column}
recs <- recs |>
  # drop original recipient recommendation status column since we now have separate columns for open/closed status
  select(-recipient_recommendation_status)
```

Now let's look at our new dataset!

```{r}
recs
```

Well, now that we've got all this cleaned up, we can write it back into a csv and save it — that way we have this version saved, and we can import it into Google Sheets if we want to.

```{r}
# write to csv
write_csv(recs, "data/clean_recs.csv")
```


# Analysis Examples: Things to Look At

### Which recipients have the most open recommendations?

```{r}
recs |>
  filter(recipient_recommendation_open_closed == "Open") |>
  group_by(recipient) |>
  count() |>
  arrange(desc(n))
```

### What are the unacceptable recommendation responses?

```{r}
recs |>
  filter(str_detect(recipient_recommendation_response_status, "Unacceptable"))
```

### How about unacceptable recommendation responses by recipient?

```{r}
recs |>
  filter(str_detect(recipient_recommendation_response_status, "Unacceptable")) |>
  group_by(recipient) |>
  count() |>
  arrange(desc(n))
```


### How many open recommendations does the NTSB have for the FRA, and how has the NTSB classified the FRA's response to those recommendations?

```{r}
recs |>
  filter(recipient == "FRA" &
         recipient_recommendation_open_closed == "Open") |>
  group_by(recipient_recommendation_response_status) |>
  count() |>
  arrange(desc(n))
```

### Other ideas:

-   How many recommendations has each government agency received, and how has the NTSB classified their response to those recommendations?

-  How many recommendations has each government agency received in each priority level, and how has the NTSB classified their response to those recommendations?

- How long does it take for recommendations to be closed, and does that vary by recipient or priority level?

- How many of the recommendations concern hazardous materials?

- Where were the accidents that led to the recommendations located?

- How many of the recommendations are on the NTSB's Most Wanted List?

- How many of the recommendations are reiterated recommendations, and how many times have they been reiterated?




## BONUS: Exploring the Full Data

### Go through these next steps on your own, if you want to try getting some additional information out of the data!

Note that the data we're dealing with here is the *summary* data from NTSB's CAROL database, which means it's missing some of the contextual information that you can see online — like a recommendation's priority level, hazmat status, etc.

The reason? The full data is in a nested JSON file that by itself is pretty unmanageable. There are ways to handle nested JSON data in R, but it can be a little tricky. You can check out the `json_loading.Rmd` file in this repo to see how I handled it, but for the sake of (your) time, I'm going to load in the cleaned version of the JSON data that I created in that file.

```{r load_cleaned_json_data}
json_data <- read_csv("bonus_json_loading/rail_safety_json_data.csv")
```

Now that we have the JSON data loaded, we can take a look at it and see what additional information it contains that we might want to analyze.

```{r view_json_data}
glimpse(json_data)
```
Wow! The JSON data contains a lot of additional information that we don't have in the summary data. Luckily, the cleaned JSON data also contains columns called `srid` and `recommendation` — just like same the`srid` and `recommendation` columns in the summary data.

Now, we're going to merge the JSON data with the summary data that we just cleaned up.  so we can use those as our keys to merge the two datasets together using a `left_join`.

```{r}
# merge json data with summary data
full_recs <- recs |>
  left_join(json_data, join_by(srid, recommendation))
```

Let's take a look!

```{r}
full_recs
```


